---
title: "Occupancy Assessment for Bats of the Pacific Northwest"
subtitle: "Preliminary Results"
author:   
  - name: Trent VanHawkins
    affiliation: On Belay Statistical Consulting
    corresponding: true
  - name: Beth Ward
    affiliation: Oregon State University-Cascades
  - name: Tom Rodhouse
    affiliation: National Park Service
    degrees: PhD
format: html
embed-resources: true
toc: true
code-fold: true
execute: 
  warning: false
code-links:
  - text: GitHub Repository
    href: https://github.com/hawktre/BatHub_Trend.git
page-layout: full
bibliography: references.bib
---

```{r}
library(tidyverse)
library(here)
library(latex2exp)
library(sf)
library(rjags)
library(coda)
library(gt)
library(kableExtra)

setwd("/Users/trentvanhawkins/Library/Mobile Documents/com~apple~CloudDocs/Documents/OnBelay/Projects/BatHub/BatHub_Trend")

# Read in Data
## Full 
dir <- "DataProcessed/results/jags/full/fits/"

filenames <- list.files(here(dir))

#read in the files into a list object
occ_jags <- lapply(filenames,function (x) readRDS(here(paste0(dir, x))))

# rename the list elements to have the correct names
names(occ_jags) <- str_split_i(filenames, "_", 1)

## Sensitivity
dir.sens <- "DataProcessed/results/jags/ORWA_Only/fits/"

filenames.sens <- list.files(here(dir.sens))

#read in the files into a list object
occ_jags_sens <- lapply(filenames.sens,function (x) readRDS(here(paste0(dir.sens, x))))

# rename the list elements to have the correct names
names(occ_jags_sens) <- str_split_i(filenames.sens, "_", 1)

## define some species we will exclude
exclude <- c("tabr")
```

# Introduction

This purpose of this report is to document an occupancy trend analysis for bat species of the Pacific Northwest. Previously, the Northwestern Bat Hub for Population Monitoring and Research (NW Bat Hub) and its partners have documented evidence of a region-wide decline for Hoary Bats between 2016 and 2018 [@rodhouse2019]. Since publication of this research, the bat hub has continued population research and monitoring under North American Bat Program (NABat) summer-monitoring protocols and has since grown to include surveys in all of Oregon, Washington, and Idaho. Here, we present the preliminary results of an updated analysis including data from 2016-2022 for all of the NABat sample units encompassed by Oregon, Washington, and Idaho.

# Methods

Methods for the current study are similar to those published in [Rodhouse et. al, 2019](https://onlinelibrary.wiley.com/doi/10.1002/ece3.5612) and @kéry2021. We will provide brief detail on important components of the data and modeling procedures but further details can be found in the literature.

## The Data

We queried data from NW Bat Hub database files for all included up to the time of this report (2016-2022). The NABat summer sampling design and its explanation are well-documented in the literature [@reichert2021; @rodhouse2017; @rodhouse2019; @udell].

Briefly, NABat uses a spatial grid covering the continental United States in 10x10 km sample units—hereto referred to as "grid cells" or simply "cells"— with an ordered sampling priority to ensure a spatially balanced sample that meets statistical modeling assumptions. The current monitoring protocol utilized by the NW Bat Hub calls for one night of acoustic monitoring at each of four spatial replicates within a cell during the summer season. Ideally, each of the spatial replicates are distributed to one of the inter-cardinal 5x5 km quadrants (NE, NW, SE, SW) of a cell, but when this is not possible multiple detectors may be deployed in a single quadrant. The current modeling strategy is also flexible to \>4 spatial replicates, but is not required. In instances where detectors were deployed for \>1 night, only the first night was subsetted for analysis.

The Mexican Free-tailed Bat (*Tadarida brasiliensis*) has been removed from this analysis due to small sample size and limited range.

## The Model

The Bayesian framework used for this analysis is thoroughly described in @rodhouse2019 and @kéry2021, but we will briefly define the model and its parameters here. The framework is broken down into two sub-models, an occupancy sub-model and a detection sub-model.

First, we have a true underlying occupancy status $Z_{it}$ at site $i$ and year $t$. This random variable can take on the value of 1 (occupied) or 0 (unoccupied). The true occupancy status is unknowable (since we cannot truly know that a site is *unoccupied* via acoustic monitoring), but we can say that it arises from a probability distribution with some probability that the site is occupied ($\psi_{it}$).

$$Z_{it} \sim \text{Bernoulli}(\psi_{it}),$$

Furthermore, we assume that the parameter $\psi_{it}$ can be partially explained by a set of environmental covariates, $\bf{x}$ (@tbl-covars), for which we will estimate an effect $\alpha$. We will also allow the probability of occupancy to vary between years by explicitly modeling the probabilities of transitioning occupancy states in years $\{2, ..., T\}$. The probability of colonization is denoted by $\gamma_{t}$ and describes the probability that a an unoccupied ($Z_{i(t-1)} = 0$) cell in year $t-1$ transitions to being occupied ($Z_{i(t)} = 1$) in year $t$. Conversely, The probability of survival is denoted by $\phi_{t}$ and describes the probability that a an occupied ($Z_{i(t-1)} = 0$) cell in year $t-1$ remains occupied ($Z_{i(t)} = 1$) in year $t$. This model parameterization is termed "fully dynamic" by @kéry2021.\
$$\text{logit}(\psi_{i1}) = \alpha_{01} + {\bf x}_i\alpha,$$ $$\text{logit}(\psi_{it}) = \gamma_{t} + {\bf x}_i\alpha + Z_{i(t-1)}\phi_{t}\text{ for }t \in \{2,\ldots,T\},$$

Finally, we have a detection sub-model which estimates the conditional probability that we detected a bat species ($p_{ijt}$) at site $i$ and replicate $j$ on year $t$ given that site $i$ was occupied at year $t$. We assume that $p_{ijt}$ arises from a similar probability distribution and may also be explained by some environmental covariates $\bf{v}$, for which we wish to estimate effects ($\bf{\beta}$).\
$$[Y_{ijt} \mid Z_{it}=1] \sim \text{Bernoulli}(p_{ijt})$$ $$\text{logit}(p_{ijt}) = \beta_0 + {\bf v}_{ijt}\beta,$$

A full summary of these parameters is provided in @tbl-params and environmental covariates in @tbl-covars. For the purpose of this report, we are most interested in reviewing the posterior probability of occupancy ($\psi_{it}$), effect estimates for occupancy environmental covariates ($\bf{\alpha}$), and effect estimates for detection environmental covariates $\bf{\beta}$. MCMC sampling was conducted in 4 chains, each run for 5000 iterations with a thinning rate of 3 and burn-in of 1000 iterations. Vague priors (Normal (mean = 0, var = 10)) priors were used universally. All models were fit using `rjags` ver. `r packageVersion("rjags")`. Code used to produce this analysis can be found in @sec-Appendix.

## Estimating Trend

The goal of this analysis is to estimate an occupancy trend for each of the species included in this analysis. There are many possible ways to accomplish this goal, but the one used in @rodhouse2019 is to define a variable

$$
\hat\lambda = \frac{\hat\psi_{(2022)}}{\hat\psi_{(2016)}}
$$

This comparison of occupancy probabilities allows estimation of relative occupancy over the duration of this study, but is no substitute for examining the posterior distributions of $\hat\psi_{it}$.

| Parameter | Description |
|----|----|
| $\psi_{it}$ | Posterior occupancy probability for site $i$ at time $t$ |
| $\alpha_{01}$ | Occupancy model intercept at year $t = 1$ |
| $\gamma_{t}$ | Probability of colonization at year $t = 2, \ldots, T$ |
| $\bf{\alpha}$ | Occupancy model effect size for environmental covariates |
| $\bf{x}$ | Occupancy model environmental covariates [@tbl-occ]. |
| $\phi_{t}$ | Autoregressive parameter allowing for different occupancy probabilities on year $t = 2,\ldots, T$ |
| $p_{ijt}$ | Conditional probability of species detection at site $i$ and replicate $j$ on year $t$ given site $i$ is occupied at year $t$ |
| $\beta_0$ | Detection model intercept |
| $\bf{v}$ | Detection model environmental covariates [@tbl-det] |
| $\bf{\beta}$ | Detection model effect sizes for environmental covariates |

: Model parameters and descriptions. {#tbl-params}

::: {#tbl-covars layout-ncol="1"}
+--------------------+----------------------------------------------------------------------+
| Name               | Description                                                          |
+====================+======================================================================+
| Forest Cover (%)   | Percentage of the sample unit that classified as forest cover (NLCD) |
+--------------------+----------------------------------------------------------------------+
| Precipitation (mm) | Mean annual precipitation (WorldClim)                                |
+--------------------+----------------------------------------------------------------------+
| Elevation (m)      | Maximum Elevation (DEM)                                              |
+--------------------+----------------------------------------------------------------------+
| Cliff Cover (%)    | Percentage of the sample unit classified as cliff cover (LandFire)   |
|                    |                                                                      |
|                    | (Anpa, Euma, Myci, Pahe)                                             |
+--------------------+----------------------------------------------------------------------+

: Environmental covariates included in the occupancy sub-model. {#tbl-occ}

| Name | Description |
|----|----|
| Minimum Temperature | Minimum temperature ($C^{\circ}$) on the night of detector deployment |
| Day Length | Day length (previous day) on the night of deployment |
| Clutter (%) | Categorical clutter (0 = 0%; 1 = 1-25%; 2 = 26-50%; 3 = 51-75%; 4 = 76-100%) |
| Waterbody | Presence of a waterbody at deployment site (1 = yes; 0 = no) |

: Environmental covariates included in the detection sub-model. {#tbl-det}

Covariates included in (A) the occupancy model, or (B) the detection model. All covariates were first standardized to mean = 0 and sd = 1 to promote computational efficiency with the exception of Clutter (centered only) and categorical Waterbody. Forest Cover and Cliff Cover were natural-log transformed before standardization.
:::

## Sensitivity Analysis

Previous studies released by the NW Bat Hub and its partners have focused on acoustic records captured in the 2016 to 2019 field seasons and, therefore, exclude Idaho. In order to compare model results from this analysis with those more comparable to previous research, we have conducted a sensitivity analysis using data from **only** Oregon and Washington. Methods for this sensitivity analysis are the same as for the primary analysis and will be presented alongside results from the full analysis.

# Results

```{r}
occ_data <- readRDS(here("DataProcessed/results/jags/occ_data.rds"))
occ_data_sens <- readRDS(here("DataProcessed/results/jags/occ_data_sens.rds"))

n_sites <- occ_data$laci$n_sites
n_sites_sens <- occ_data_sens$laci$n_sites
```

A total of `r n_sites` independent cells were included across the seven year period from 2016 to 2022 in the primary analysis and `r n_sites_sens` in the sensitivity analysis (@tbl-surveys; @fig-samples).

```{r}
#| label: tbl-surveys
#| tbl-cap: "Sample units surveyed by year"

occ_map <- readRDS(here("DataProcessed/results/stan/full/occ_map.rds"))

occ_map$anpa$means_map_long %>% 
  st_drop_geometry() %>% 
  filter(sampled == 1) %>% 
  group_by(years) %>% 
  reframe("Sample Units" = sum(sampled)) %>% 
  rename("Year" = years) %>% 
  gt() %>% 
  tab_options(table.width = pct(50)) %>% 
  cols_align("left")
```

Partners at in Idaho joined the NABat effort coordinated by the NW Bat Hub during the 2020 season, and sampling efforts are reflected in @fig-samples.

```{r fig.height=8, fig.width=8}
#| label: fig-samples
#| fig-cap: "Map of NABat grid cells (10x10 km) sampled by year."
occ_map$anpa$means_map_long %>%
  mutate(sampled = factor(sampled, labels = c("No", "Yes"))) %>% 
  ggplot()+
  geom_sf(aes(fill = factor(sampled)))+
  facet_wrap(~years)+
  scale_fill_discrete(type = c("transparent", "blue"))+
  labs(fill = "Sampled")+
  theme_bw()+
  theme(axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank(),
        legend.position = "bottom")
  
  
```

## Modeling Results

All models satisfied MCMC convergence criteria with nominal $\hat{R}$ (\<1.1) and Effective Sample Size (\>500) values. Adequate chain mixing was assessed by visual inspection of traceplots.

### Covariates

@fig-alphas gives summaries of the posterior distributions for each environmental covariate from the occupancy sub-model, for each species included in the analysis.

```{r, fig.height=8, fig.width=8}
#| label: fig-alphas
#| fig-cap: "Summary of posterior distributions for occupancy model covariates, by species. Point locations represent the posterior means, colored regions represent the 50% credible interval, and black lines represent the 95% credible interval. Outputs provided on the log-odds scale."

# Extract mcmc draws and convert to dataframe for summarization and plotting (outputs all.jags which contains all mcmc draws of all params for all species)
mcmc_extract <- function(occ_jags){
  for(i in 1:length(occ_jags)){
    cur.fit <- occ_jags[[i]]
    spp <- names(occ_jags)[i]
  
    for(j in 1:length(cur.fit)){
      tmp <- cur.fit[[j]] %>% 
        as.data.frame() %>% 
        mutate(iter = seq(1:nrow(cur.fit[[j]])),
               chain = j,
               spp = spp)
      if(j == 1){dat <- tmp}
      else{dat <- bind_rows(dat, tmp)}
    }
    if(i == 1){all.jags <- dat}
    else{all.jags <- bind_rows(all.jags, dat)}
  }
  return(all.jags)
}
## Use our function to extract our data
all.jags <- bind_rows(mcmc_extract(occ_jags) %>% mutate(analysis = "OR|WA|ID"), mcmc_extract(occ_jags_sens) %>% mutate(analysis = "OR|WA Only"))


## Summarise parameter estimates
param_summ <- all.jags %>%
  select(dplyr::contains('alpha') | dplyr::contains('beta'), spp, analysis) %>% 
  rename("Min. Temp" = beta1,
         "Daylight" = beta2,
         "Clutter 0" = beta3,
         "Clutter 1" = beta4,
         "Clutter 2" = beta5,
         "Clutter 3" = beta6,
         "Water" = beta7,
         "Log(Forest Cover)" = `alphas[1]`,
         "Precipitation" = `alphas[2]`,
         "Elevation" = `alphas[3]`,
         "Log(Cliff Cover)" = `alphas[4]`) %>% 
  pivot_longer(cols = -c(spp, analysis), names_to = "param") %>% 
  group_by(spp, analysis, param) %>% 
  summarise(mean = mean(value),
            q2.5 = quantile(value, probs = c(0.025), na.rm = T),
            q25 = quantile(value, probs = c(0.25), na.rm = T),
            q50 = quantile(value, probs = c(0.5), na.rm = T),
            q75 = quantile(value, probs = c(0.75), na.rm = T),
            q97.5 = quantile(value, probs = c(0.975), na.rm = T)) %>% 
  ungroup() %>% 
  mutate(type = if_else(param %in% c("Log(Forest Cover)", "Precipitation", "Elevation", "Log(Cliff Cover)", "alpha01"), "Occurrence", "Detection")) 
  


##Plot alphas
param_summ %>% 
  filter(type == "Occurrence") %>% 
ggplot(aes(x = spp, y = mean, group = analysis))+
  geom_errorbar(aes(ymin = q2.5, ymax = q97.5), width = 0, size = 0.75,
                position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = q25, ymax = q75, color = analysis), width = 0,
                size = 1.5, position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  theme_bw() +
  facet_grid(param ~ ., scales = 'free') +
  geom_hline(yintercept = 0, lty = 2) +
  xlab('Species') + 
  ylab('Posterior Distribution (Log-Odds)') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10)) +
  ggtitle('Comparing occupancy coefficients (Mean & 95% CI)',
          'Multi-season, single-species model')+
  labs(color = "Legend")
  
```

@fig-alphas shows stable estimates between the full and sensitivity analyses. Estimates are reported on the log-odds scale and can be generally be interpreted as "a change of 1 standard deviation in X is associated with an increase of Y in the log-odds of occupancy".

@fig-betas gives summaries of the posterior distributions for each environmental covariate from the detection sub-model, for each species included in the analysis.

```{r fig.height=8, fig.width=8}
#| label: fig-betas
#| fig-cap: "Summary of posterior distributions for detection model covariates, by species. Point locations represent the posterior means, colored regions represent the 50% credible interval, and black lines represent the 95% credible interval. Outputs provided on the log-odds scale."
param_summ %>% 
  filter(type == "Detection") %>% 
ggplot(aes(x = spp, y = mean, group = analysis))+
  geom_errorbar(aes(ymin = q2.5, ymax = q97.5), width = 0, size = 0.75,
                position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = q25, ymax = q75, color = analysis), width = 0,
                size = 1.5, position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  theme_bw() +
  facet_grid(param ~ ., scales = 'free') +
  geom_hline(yintercept = 0, lty = 2) +
  xlab('Species') + 
  ylab('Posterior Distribution (Log-Odds)') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10)) +
  ggtitle('Comparing detection coefficients (Mean & 95% CI)',
          'Multi-season, single-species model')+
  labs(color = "Legend")
```

@fig-betas again shows stable estimates for model coefficients across species. These results may also be interpreted as "a change of 1 standard deviation in X is associated with an increase of Y in the log-odds of detection, given that the species is present."

### Occupancy

@fig-psi gives a summary of the posterior distribution of estimated occupancy probabilities across all predicted grid cells in the three-state region of Oregon, Washington, and Idaho. Again, these are preliminary results and are subject to change. Exact estimates and credible intervals are available in @tbl-psi and @tbl-psi-sens for sensitivity results.

```{r, fig.height=20}
#| label: fig-psi
#| fig-cap: "Region-wide summary of the posterior distribution of $\\hat{\\psi_t}$ along with 50% credible interval (colored region) and 95% credible interval (black line)."
psi_summ <- all.jags %>% 
  select(contains("psi"), spp, analysis) %>% 
  pivot_longer(-c(spp, analysis)) %>% 
  mutate(year = case_when(str_detect(name, '1')~2016,
                          str_detect(name, '2')~2017,
                          str_detect(name, '3')~2018,
                          str_detect(name, '4')~2019,
                          str_detect(name, '5')~2020,
                          str_detect(name, '6')~2021,
                          str_detect(name, '7')~2022),
         type = if_else(str_detect(name, 'avg'), 'Average', 'Hat')) %>% 
  group_by(spp, analysis, type, year) %>% 
  summarise(mean = mean(value),
            q2.5 = quantile(value, probs = c(0.025), na.rm = T),
            q25 = quantile(value, probs = c(0.25), na.rm = T),
            q50 = quantile(value, probs = c(0.5), na.rm = T),
            q75 = quantile(value, probs = c(0.75), na.rm = T),
            q97.5 = quantile(value, probs = c(0.975), na.rm = T)) %>% 
  ungroup()

psi_summ %>% 
  filter(type == "Hat") %>% 
  ggplot(aes(x = as.factor(year), y = mean, group = analysis))+
  geom_errorbar(aes(ymin = q2.5, ymax = q97.5), width = 0, size = 0.75,
                position = position_dodge(width = 0.25)) +
  geom_errorbar(aes(ymin = q25, ymax = q75, color = analysis), width = 0,
                size = 1.5, position = position_dodge(width = 0.25)) +
  geom_point(position = position_dodge(width = 0.25)) +
  theme_bw() +
  facet_grid(spp ~ ., scales = 'free') +
  xlab('Year') + 
  ylab(expression('Posterior Distribution' ~ hat(psi)[t])) +
  ylim(c(0,1))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10)) +
  ggtitle('Occurrence Probabilities Over Time',
          'Multi-season, single-species model')+
  labs(color = "Legend") 
```

## Trend Estimation

As mentioned previously, we define $\lambda = \frac{\hat\psi_{2022}}{\hat\psi_{2016}}$. This provides a relative comparison of trend over the endpoints of our study period.

```{r fig.width=8}
#| label: fig-lambda1
#| fig-cap: "$\\hat{\\lambda}$ estimated for each species."
lambda_summ <- all.jags %>% 
  select(contains("lam.tot"), spp, analysis) %>% 
  pivot_longer(-c(spp, analysis)) %>% 
  group_by(spp, analysis) %>% 
  summarise(mean = mean(value),
            q2.5 = quantile(value, probs = c(0.025), na.rm = T),
            q25 = quantile(value, probs = c(0.25), na.rm = T),
            q50 = quantile(value, probs = c(0.5), na.rm = T),
            q75 = quantile(value, probs = c(0.75), na.rm = T),
            q97.5 = quantile(value, probs = c(0.975), na.rm = T)) %>% 
  ungroup()

lambda_summ %>% 
  ggplot(aes(x = spp, y = mean, group = analysis))+
  geom_errorbar(aes(ymin = q2.5, ymax = q97.5), width = 0, size = 0.75,
                position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = q25, ymax = q75, color = analysis), width = 0,
                size = 1.5, position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  theme_bw() +
  xlab('Year') + 
  ylab(expression('Posterior Distribution' ~ hat(lambda))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10)) +
  ggtitle('Relative Occupancy Probability in 2022 vs. 2016')+
  labs(color = "Legend") 
```

Estimates for *coto* and *euma* in @fig-lambda1 appear to be extremely more likely to occur in 2022 than 2016. This is most likely due to the fact that they have very low occurrence probabilities in 2016, so even small gains in improvement would lead to large values of $\lambda$. We review the figure without those species in @fig-lambda2.

```{r fig.width=8}
#| label: fig-lambda2
#| fig-cap: "$\\hat{\\lambda}$ estimated for each species with coto and euma removed to better visualize results. The dashed horizontal line indicates $\\hat{\\lambda} = 1$. 95% credible intervals falling strictly below this line indicate a species that is less likely to occur in 2022 than 2016."

lambda_summ %>% filter(!spp %in% c("coto", "euma")) %>% 
  ggplot(aes(x = spp, y = mean, group = analysis))+
  geom_errorbar(aes(ymin = q2.5, ymax = q97.5), width = 0, size = 0.75,
                position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = q25, ymax = q75, color = analysis), width = 0,
                size = 1.5, position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  theme_bw() +
  geom_hline(yintercept = 1, lty = "dashed")+
  xlab('Year') + 
  ylab(expression('Posterior Distribution' ~ hat(lambda))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10)) +
  ggtitle('Relative Occupancy Probability in 2022 vs. 2016')+
  labs(color = "Legend") 
```

We see a decline in relative occurrence probabilities ($\hat\lambda < 1$) between 2016 and 2022 for EPFU, MYCA, MYEV, MYTH, and MYVO. Evidence is particularly strong for EPFU, MYTH, and MYVO.

### Occupancy maps

Maps of the posterior predictions for all species displayed in @fig-psi are included with this report and visualize the mean and 95% credible intervals of the posterior distribution of $\hat{\psi_t}$. @fig-psimap gives an example of the map for mean posterior occupancy probabilities and @fig-psici displays uncertainty around those means.

```{r fig.width=8}
#| label: fig-psimap
#| fig-cap: "Mean posterior occupancy probabilities visualized for the small-footed bat. Cool colors indicate low probability of occurrence on year *t* and warm colors indicate a high probability."

psi.post.map <- readRDS(here("DataProcessed/results/jags/psi_post_map.rds"))

psi.post.map %>% 
  filter(spp == "myci") %>% 
  ggplot()+
    geom_sf(aes(fill = mean), size = 0.1, lwd = 0.05) +
    facet_wrap(. ~ year, ncol = 4) +
    viridis::scale_fill_viridis(limits = c(0, 1)) +
    theme(axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "bottom")+
    ggtitle('MYCI Posterior Mean Occupancy',
            'Multi-season, single-species model')+
  labs(fill = "Posterior Mean")
```

```{r fig.width=8}
#| label: fig-psici
#| fig-cap: "95% credible interval of posterior occupancy probabilities visualized for the small-footed bat. Warm colors indicate higher uncertainty."
psi.post.map %>% 
  filter(spp == "myci") %>% 
    ggplot()+
    geom_sf(aes(fill = width), size = 0.1, lwd = 0.05) +
    facet_wrap(. ~ year, ncol = 4) +
    viridis::scale_fill_viridis(limits = c(0, 1), option = "magma") +
    theme(axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "bottom")+
    ggtitle('MYCI Posterior 95% Credible Interval (Width)',
            'Multi-season, single-species model')+
  labs(fill = "95% CI")
```

# Discussion

This report represents a significant update to monitoring of regional trends for bat species of the pacific northwest. We used the fully dynamic occupancy model proposed by @kéry2021 and implemented by @rodhouse2019 to summarize occupancy probabilities in the three-state region of Oregon, Washington, and Idaho for each of 14 bat species. We note a measurable decline in occupancy probabilities for *Eptesicus fuscus*, *Myotis volans*, and *Myotis thysanodes* across the three state region of Oregon, Washington, and Idaho.

Dynamic occupancy models have proven to be a powerful tool region-wide population monitoring as they can directly model the latent ecological processes of occupancy, colonization, survival, and extinction and provide optimal estimates for occupancy probabilities ($\hat\psi$) across years. Future work may want to focus on developing more precise tools for linear and non-linear estimates of the effect of time (in this case year) on the log-odds of occupancy, as these parameters could provide a clean and clear interpretation of these models.

# Acknowledgements

The Northwest Bat Hub thanks its contributing partners for their continued support to better understand regional status and trends of bats in the Pacific Northwest.

# Appendix {#sec-Appendix}

## Figures & Tables

```{r}
#| label: tbl-psi
#| tbl-cap: "Region-wide summary of the posterior distribution of $\\hat{\\psi_t}$ with credible intervals for the analysis including Oregon, Washington, and Idaho"
library(readxl)
wbwg <- read_xlsx("/Users/trentvanhawkins/Library/Mobile Documents/com~apple~CloudDocs/Documents/OnBelay/Projects/BatHub/BatHub_SDM/Background/SpeciesNatHistMatrix.xlsx", sheet = 2)

name.match <- wbwg %>% select(fourlettername, SPECIES) %>% drop_na()
psi_summ <- psi_summ %>% left_join(name.match, by = c("spp" = 'fourlettername'))
psi_summ %>% 
  filter(type == "Hat", analysis == "OR|WA|ID") %>% 
  select(SPECIES, year, mean, q2.5, q25, q75, q97.5) %>% 
  mutate(year = as.factor(year)) %>% 
  gt(groupname_col = "SPECIES") %>%
  cols_label("mean" ~ "Mean",
             "year" ~ "Year",
             "q2.5" ~ "2.5%",
             "q25" ~ "25%",
             "q75" ~ "75%",
             "q97.5" ~ "97.5%") %>% 
  tab_spanner(label = "Quantile", columns = 4:7) %>% 
  fmt_number(decimals = 2) %>% 
  tab_options(row_group.as_column = T)
```

```{r}
#| label: tbl-psi-sens
#| tbl-cap: "Region-wide summary of the posterior distribution of $\\hat{\\psi_t}$ with credible intervals for the analysis including Oregon and Washington only"

psi_summ %>% 
  filter(type == "Hat", analysis == "OR|WA Only") %>% 
  select(SPECIES, year, mean, q2.5, q25, q75, q97.5) %>% 
  mutate(year = as.factor(year)) %>% 
  gt(groupname_col = "SPECIES") %>%
  cols_label("mean" ~ "Mean",
             "year" ~ "Year",
             "q2.5" ~ "2.5%",
             "q25" ~ "25%",
             "q75" ~ "75%",
             "q97.5" ~ "97.5%") %>% 
  tab_spanner(label = "Quantile", columns = 4:7) %>% 
  fmt_number(decimals = 2) %>% 
  tab_options(row_group.as_column = T)
```

## JAGS Code

```{jags}

model {
  # Specify priors
  ## Dynamics Parameters
  for (t in 1:(n_years-1)){
    phi[t] ~ dnorm(0, 0.1) #survival
    gamma[t] ~ dnorm(0, 0.1) #colonization
  }

  ## Occurrence-level Parameters
  alpha01 ~ dnorm(0, 0.1) #Year 1 occurrence model intercept

  for(n in 1:n_xcovs){
    alphas[n] ~ dnorm(0, 0.1) #Occurrence covariate intercepts
  }

  ## Detection-level Parameters
  beta0 ~ dnorm(0,0.1) #beta0
  beta1 ~ dnorm(0,0.1) #tmin covariate
  beta2 ~ dnorm(0,0.1) #daylight covariate
  beta3 ~ dnorm(0,0.1) #clut0 covariate
  beta4 ~ dnorm(0,0.1) #Clut1 covariate
  beta5 ~ dnorm(0,0.1) #Clut2 covariate
  beta6 ~ dnorm(0,0.1) #Clut3 covariate (ref is 4)
  beta7 ~ dnorm(0,0.1) #water indicator covariate

  # Fit our Occurrence-level submodel

  ## First Year
  logit_psi[1:n_sites, 1] <- alpha01 + xmat %*% alphas

  ## Following years
  for (t in 1 : (n_years - 1)) {
      logit_psi[1:n_sites, t + 1] <- gamma[t] + xmat %*% alphas #Get gamma colonization term as intercept in following years
  }

  # Ecological submodel: Define state conditional on parameters
  for (i in 1:n_sites){

    z[i,1] ~ dbern(psi[i,1]) #latent occupancy state in year 1 ~ bernoulli with prob based on our covariates and alpha01
    psi[i,1] <- ilogit(logit_psi[i,1])

    #Detection Model Year 1
    for(j in 1:n_visits[i, 1]){
      dets[i, j, 1] ~ dbern(z[i, 1] * p[i, j, 1])
      p[i, j, 1] <- ilogit(beta0 +
        beta1 * tmin[i, j, 1] +
        beta2 * dayl[i, j, 1] +
        beta3 * clut0[i, j, 1] +
        beta4 * clut1[i, j, 1] +
        beta5 * clut2[i, j, 1] +
        beta6 * clut3[i, j, 1] +
        beta7 * wind[i, j, 1])
    }
    
    #Occurrence Probs and Latent state for years 2 to n_years
    for (t in 2:n_years){
      z[i,t] ~ dbern(psi[i,t])
      psi[i,t] <- ilogit(logit_psi[i, t] + z[i, (t-1)] * phi[t-1]) 
    
      #Detection model for years 2 to n_years
      for(j in 1:n_visits[i, t]){
        dets[i, j, t] ~ dbern(z[i, t] * p[i, j, t])
        p[i, j, t] <- ilogit(beta0 + 
        beta1 * tmin[i, j, t] +
        beta2 * dayl[i, j, t] +
        beta3 * clut0[i, j, t] +
        beta4 * clut1[i, j, t] +
        beta5 * clut2[i, j, t] +
        beta6 * clut3[i, j, t] +
        beta7 * wind[i, j, t])
      }
    }
  }
  # Derived parameters
  # lambda measurements (trend)
  lam.tot<-psi.hat[7]/psi.hat[1] 
  lam.avg<-mean(psi.hat[])
  lam.tot.avg<-avg.psi[7]/avg.psi[1]
  lam.avg.avg<-mean(avg.psi[])
  
  #Original estimate for psi.hat (just the intercept of our original model)
  psi.hat[1]<-ilogit(alpha01)
  
  #Psi estimates for following years
  for (t in 2:n_years){
    col[t]<-ilogit(gamma[t-1])
    surv[t]<-ilogit(gamma[t-1]+phi[t-1])
    psi.hat[t]<-psi.hat[t-1]*surv[t]+(1-psi.hat[t-1])*col[t]
    turnover[t]<-((1-psi.hat[t-1])*col[t])/((1-psi.hat[t-1])*col[t]+surv[t]*psi.hat[t-1])
    }
  #Avg psi at year 1
  avg.psi[1] <- mean(psi[ ,1])
  
  #Avg psi at year t
  for(t in 2:n_years){
    avg.psi[t] <- mean(psi[,t])
  }
    
  for (t in 1:n_years){
    p.hat[t]<-ilogit(beta0)
    }
}

```

# References

::: {#refs}
:::
